A space-time skew-$t$ model for threshold exceedances

### 2015-09-17 LaTeX/skewnormal.tex
setup biometrics style

\documentclass[11pt]{article}
\usepackage{amssymb, amsthm, amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage{bm}
\usepackage{verbatim}
\usepackage{lineno}
\usepackage{times}
\usepackage{soul}
\usepackage{color}
\usepackage{enumitem}
\usepackage{setspace}

@@

\usepackage[left=1in,top=1in,right=1in]{geometry}
\pdfpageheight 11in
\pdfpagewidth 8.5in
\linespread{2.0}
\input{mycommands.sty}

\begin{document}\linenumbers
\begin{center}
{\Large {\bf A space-time skew-$t$ model for threshold exceedances}}\\

{\large Samuel A Morris\footnote[1]{North Carolina State University}, Brian J Reich\footnotemark[1]{}, Emeric Thibaud\footnote[2]{Colorado State University}, and Daniel Cooley\footnotemark[2]{}}

\today
\end{center}

\begin{abstract}
\singlespacing
To assess the compliance of air quality regulations, the Environmental Protection Agency (EPA) must know if a site exceeds a pre-specified threshold.
In the case of ozone, the threshold for compliance is fixed at 75 parts per billion, which is high, but not extreme at all locations.
We present a new method based on the spatial skew-$t$ process.
Our method incorporates a random partition to permit long-distance asymptotic independence while allowing for sites that are near one another to be asymptotically dependent, and we incorporate thresholding to allow the tails of the data to speak for themselves.
We also introduce a transformed AR(1) time-series to allow for temporal dependence.
Finally, our model allows for high-dimensional Bayesian inference that is comparable in speed to traditional geostatistical methods for large datasets.
We apply our method to an ozone analysis for July 2005, and find that our model improves over both Gaussian and max-stable methods in terms of predicting exceedances over a fixed threshold.

{\bf Key words}: Skew-$t$, random partition, MCMC, extreme value analysis, spatio-temporal
\end{abstract}

\newpage
\section{Introduction}\label{s:intro}
% In many climatological applications, researchers are interested in learning about the average behavior of different climate variables (e.g. ozone, temperature, rainfall).
Epidemiological studies have linked air quality to public health concerns regarding morbidity and mortality \citep{Samet2000}.
As a result, the Environmental Protection Agency (EPA) has developed a set of standards to help reduce air pollution thereby improving air quality.
Our study is motivated by an air pollution application where the focus is not on the average behavior, but instead the behavior over a fixed threshold determined by government regulation.
More specifically, we consider the case of compliance for ozone.
A site is said to be in compliance if the fourth highest daily maximum 8-hour concentration averaged over 3 years does not exceed 75 parts per billion (ppb).
Figure \ref{fig:ozone-10jul} shows the ozone levels from July 10, 2005, at 1089 stations across the United States.
We see a large area above the compliance level in the midwest covering Ohio, Indiana, Illinois, and parts of the surrounding states.

@@

\subsection{Extremal dependence}\label{s:extdep}
Our interest lies in spatial dependence in the tail of the skew-$t$ process.
One measure of extremal dependence is the $\chi$ statistic \citep{Coles1999}.
For a stationary and isotropic spatial process, the $\chi$ statistic for locations $\bs$ and $\bt$ separated by distance $h = || \bs - \bt ||$ with identical marginal distributions is
\begin{align}
  \chi(h) = \lim_{c \rightarrow c^*} \Pr[Y(\bs) > c | Y(\bt) > c]
\end{align}
where $c^*$ is the upper limit of the support of $Y$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the distance $h$, so they are not suitable for modeling asymptotically dependent extremes.
Unlike the Gaussian process, the skew-$t$ process is asymptotically dependent (the explicit expression for $\chi(h)$ is given in Appendix \ref{a:skewt}).
However, one problem with the spatial skew-$t$ process is that $\lim_{h \rightarrow \infty} \chi(h) > 0$.
This occurs because all observations, both near and far, share the same $z$ and $\sigma$ terms.
Therefore, this long-range dependence feature of the skew-$t$ process is not ideal for spatial analysis of large geographic regions where we expect only local spatial dependence.
We propose a solution to this in Section \ref{s:part}.

@@

The partitioning model remove long-range dependence. Conditional on knots $\bw_1, \ldots, \bw_K$, the $\chi$ statistic for two sites $\bs$ and $\bt$ in partitions $k_s$ and $k_t$ respectively is
\begin{align}
  \chi(h) &= I(k_s = k_t) \chi_{\text{skew-}t}(h) + I(k_s \neq k_t) \chi_{\text{Gaus}}(h) \nonumber \\
         &= I(k_s = k_t) \chi_{\text{skew-}t}(h)
\end{align}
where $I(\cdot)$ is an indicator function, $\chi_{\text{skew-}t}(h)$ is the $\chi$ statistic for a skew-$t$ process given in (\ref{eq:chiskew-t}), $\chi_{\text{Gaus}}(h)$ is the $\chi$ statistic for a Gaussian process, and $h = ||\bs - \bt||$.
Therefore, sites in different subregions are asymptotically independent because $\chi_{\text{Gaus}}(h) = 0$ for all $h$.
Marginally, over the knots, $\chi(h) = \pi(h) \chi_{\text{skew}-t}(h)$, where $\pi(h) = \Pr(k_s = k_t)$ is the probability that two sites separated by distance $h$ are in the same partition.
In Appendix \ref{a:proofsamepartition}, we show that $\lim_{h \rightarrow \infty} \pi(h) = 0$, implying $\lim_{h \rightarrow \infty} \chi(h) = 0$.
In Figure \ref{fig:chi}, we give $\chi(h)$ for $K = 1, 3, 5, 10$ partitions for a skew-$t$ distribution with $\alpha = 10$, and 3 degrees of freedom.
In Figure \ref{fig:chi} we estimate $\pi(h)$ through simulation.
% To estimate $\pi(h)$, we generate 500 sites uniformly over the unit-square.
% We then randomly generate 400 different sets of partitions using $K = 3$, $5$, and $10$.
% For each set of knots, we take $\pi(h)$ to be the proportion of sites in the same partition that are separated by distance $h$.
% This plot demonstrates how partitioning helps to reduce extremal dependence as $h$ increases.

@@

\subsection{Results}\label{s:simresults}
We compared the Brier scores for exceeding 4 different thresholds for each dataset.
The thresholds used for the Brier scores are extreme quantiles from the simulated data for $q(0.90)$, $q(0.95)$, $q(0.98)$, and $q(0.99)$.
Figure \ref{fig:simbrierscores} gives the Brier score relative to the Brier score for the Gaussian method calculated as
\begin{align}
  \text{BS}_{\text{rel}} = \frac{\text{BS}_{\text{method}}}{\text{BS}_{\text{Gaussian}}}.
\end{align}
We analyzed the results for the simulation study using a Friedman test at $\alpha = 0.05$ to see if at least one method had a significantly different Brier score.
For Friedman tests that came back with a significant p-value, we conducted a Wilcoxon-Nemenyi-McDonald-Thompson test to see which of the methods had different results.
The full results for the Wilcoxon-Nemenyi-McDonald-Thompson tests are given in Appendix \ref{a:pdiffs}.

@@

@@

\begin{singlespace}
\bibliographystyle{rss}
\bibliography{library}
\end{singlespace}

>>>>

% \documentclass[11pt]{article}
\documentclass[useAMS,usenatbib,referee]{biom}
% \usepackage{amssymb, amsthm, amsmath}
\usepackage{amssymb, amsmath}
\usepackage{bm}
\usepackage{graphicx}
% \usepackage[authoryear]{natbib}
\usepackage{natbib}
\usepackage{bm}
\usepackage{verbatim}
\usepackage{lineno}
\usepackage{times}
\usepackage{soul}
\usepackage{color}
\usepackage{enumitem}
\usepackage{setspace}

@@

% \usepackage[left=1in,top=1in,right=1in]{geometry}
% \pdfpageheight 11in
% \pdfpagewidth 8.5in
% \linespread{2.0}
\input{mycommands.sty}

\title[A Space-time Skew-$t$ Model for Threshold Exceedances]{A space-time skew-$t$ model for threshold exceedances}
\author
{Samuel A Morris$^{1,*}$\email{samorris@ncsu.edu},
Brian J Reich$^{1}$,
Emeric Thibaud$^{2}$, and
Daniel Cooley$^{2}$\\
$^{1}$Department of Statistics, North Carolina State University, Raleigh, North Carolina, U.S.A. \\
$^{2}$Department of Statistics, Colorado State University, Fort Collins, Colorado, U.S.A.}

\begin{document} %\linenumbers
% \begin{center}
% {\Large {\bf A space-time skew-$t$ model for threshold exceedances}}\\

% {\large Samuel A Morris\footnote[1]{North Carolina State University}, Brian J Reich\footnotemark[1]{}, Emeric Thibaud\footnote[2]{Colorado State University}, and Daniel Cooley\footnotemark[2]{}}

% \today
% \end{center}

\pagerange{\pageref{firstpage}--\pageref{lastpage}}

\label{firstpage}

\begin{abstract}
To assess the compliance of air quality regulations, the Environmental Protection Agency (EPA) must know if a site exceeds a pre-specified threshold.
In the case of ozone, the threshold for compliance is fixed at 75 parts per billion, which is high, but not extreme at all locations.
We present a new method based on the spatial skew-$t$ process.
Our method incorporates a random partition to permit long-distance asymptotic independence while allowing for sites that are near one another to be asymptotically dependent, and we incorporate thresholding to allow the tails of the data to speak for themselves.
We also introduce a transformed AR(1) time-series to allow for temporal dependence.
Finally, our model allows for high-dimensional Bayesian inference that is comparable in speed to traditional geostatistical methods for large datasets.
We apply our method to an ozone analysis for July 2005, and find that our model improves over both Gaussian and max-stable methods in terms of predicting exceedances over a fixed threshold.
\end{abstract}

\begin{keywords}
Skew-$t$, random partition, MCMC, extreme value analysis, spatio-temporal
\end{keywords}

\maketitle

\section{Introduction}\label{s:intro}
% In many climatological applications, researchers are interested in learning about the average behavior of different climate variables (e.g. ozone, temperature, rainfall).
Epidemiological studies have linked air quality to public health concerns regarding morbidity and mortality \citep{Samet2000}.
As a result, the Environmental Protection Agency (EPA) has developed a set of standards to help reduce air pollution thereby improving air quality.
Our study is motivated by an air pollution application where the focus is not on the average behavior, but instead the behavior over a fixed threshold determined by government regulation.
More specifically, we consider the case of compliance for ozone.
A site is said to be in compliance if the fourth highest daily maximum 8-hour concentration averaged over 3 years does not exceed 75 parts per billion (ppb).
Figure \ref{fig:ozone-10jul} shows the ozone levels from July 10, 2005, at 1089 stations across the United States.
We see a large area above the compliance level in the midwest covering Ohio, Indiana, Illinois, and parts of the surrounding states.

@@

\subsection{Extremal dependence}\label{s:extdep}
Our interest lies in spatial dependence in the tail of the skew-$t$ process.
One measure of extremal dependence is the $\chi$ statistic \citep{Coles1999}.
For a stationary and isotropic spatial process, the $\chi$ statistic for locations $\bs$ and $\bt$ separated by distance $h = || \bs - \bt ||$ with identical marginal distributions is
\begin{align}
  \chi(h) = \lim_{c \rightarrow c^*} \Pr[Y(\bs) > c | Y(\bt) > c]
\end{align}
where $c^*$ is the upper limit of the support of $Y$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the distance $h$, so they are not suitable for modeling asymptotically dependent extremes.
Unlike the Gaussian process, the skew-$t$ process is asymptotically dependent (the explicit expression for $\chi(h)$ is given in \ref{a:skewt}).
However, one problem with the spatial skew-$t$ process is that $\lim_{h \rightarrow \infty} \chi(h) > 0$.
This occurs because all observations, both near and far, share the same $z$ and $\sigma$ terms.
Therefore, this long-range dependence feature of the skew-$t$ process is not ideal for spatial analysis of large geographic regions where we expect only local spatial dependence.
We propose a solution to this in Section \ref{s:part}.

@@

The partitioning model remove long-range dependence. Conditional on knots $\bw_1, \ldots, \bw_K$, the $\chi$ statistic for two sites $\bs$ and $\bt$ in partitions $k_s$ and $k_t$ respectively is
\begin{align}
  \chi(h) &= I(k_s = k_t) \chi_{\text{skew-}t}(h) + I(k_s \neq k_t) \chi_{\text{Gaus}}(h) \nonumber \\
         &= I(k_s = k_t) \chi_{\text{skew-}t}(h)
\end{align}
where $I(\cdot)$ is an indicator function, $\chi_{\text{skew-}t}(h)$ is the $\chi$ statistic for a skew-$t$ process given in (\ref{eq:chiskew-t}), $\chi_{\text{Gaus}}(h)$ is the $\chi$ statistic for a Gaussian process, and $h = ||\bs - \bt||$.
Therefore, sites in different subregions are asymptotically independent because $\chi_{\text{Gaus}}(h) = 0$ for all $h$.
Marginally, over the knots, $\chi(h) = \pi(h) \chi_{\text{skew}-t}(h)$, where $\pi(h) = \Pr(k_s = k_t)$ is the probability that two sites separated by distance $h$ are in the same partition.
In \ref{a:proofsamepartition}, we show that $\lim_{h \rightarrow \infty} \pi(h) = 0$, implying $\lim_{h \rightarrow \infty} \chi(h) = 0$.
In Figure \ref{fig:chi}, we give $\chi(h)$ for $K = 1, 3, 5, 10$ partitions for a skew-$t$ distribution with $\alpha = 10$, and 3 degrees of freedom.
In Figure \ref{fig:chi} we estimate $\pi(h)$ through simulation.
% To estimate $\pi(h)$, we generate 500 sites uniformly over the unit-square.
% We then randomly generate 400 different sets of partitions using $K = 3$, $5$, and $10$.
% For each set of knots, we take $\pi(h)$ to be the proportion of sites in the same partition that are separated by distance $h$.
% This plot demonstrates how partitioning helps to reduce extremal dependence as $h$ increases.

@@

\subsection{Results}\label{s:simresults}
We compared the Brier scores for exceeding 4 different thresholds for each dataset.
The thresholds used for the Brier scores are extreme quantiles from the simulated data for $q(0.90)$, $q(0.95)$, $q(0.98)$, and $q(0.99)$.
Figure \ref{fig:simbrierscores} gives the Brier score relative to the Brier score for the Gaussian method calculated as
\begin{align}
  \text{BS}_{\text{rel}} = \frac{\text{BS}_{\text{method}}}{\text{BS}_{\text{Gaussian}}}.
\end{align}
We analyzed the results for the simulation study using a Friedman test at $\alpha = 0.05$ to see if at least one method had a significantly different Brier score.
For Friedman tests that came back with a significant p-value, we conducted a Wilcoxon-Nemenyi-McDonald-Thompson test to see which of the methods had different results.
The full results for the Wilcoxon-Nemenyi-McDonald-Thompson tests are given in \ref{a:pdiffs}.

@@

\backmatter

@@

\section*{Supplementary Materials}

\bibliographystyle{biom}
\bibliography{library}

\label{lastpage}
